---
title: "readme"
author: "Pietro Franceschi"
date: "`r Sys.Date()`"
output: github_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Introduction

This md file documents the data analysis workflow used to produce the results presented in the paper *Correlation between tree-ring series as a dendroprovenancing evaluation tool* by M. Bernabei and P. Franceschi.

## Includes

This is the set of libraries required to perform the analyses

```{r message=FALSE, warning=FALSE}
library(dplR)                
library(tidyverse)
library(sf)
library(knitr)
library(magrittr)
library(quantreg)
library(colorspace)
library(jtools)
library(broom)
```

## Data

11,972 individual, raw and georeferenced tree ring series, referring to three important Alpine species, Norway spruce (Picea abies Karst., PCAB), larch (Larix decidua Mill., LADE) and silver fir (Abies alba Mill., ABAL), were downloaded from the ITRDB site ([International Tree Ring Data Bank, United States’ National Oceanic and Atmospheric Administration Paleoclimatology Program and World Data Center for Paleoclimatology](https://www.ncei.noaa.gov/products/paleoclimatology/tree-ring) (Guiterman et al., 2024).

The data were organised into a complete tibble which is made available in the repository (`full_series.RData`).

```{r}
## load the dataset
load("full_series.RData")

## print the column names
colnames(dat)
```
* `meta`: the metainformation present in the NOAA db.
* `study_code`: the identifier of the study
* `specie`: the tree specie code
* `geotype`: the type of geographical object (here always a POINT)
* `minelev`,`maxelev`: the altitude boundaries of the series, if present
* `geometry`: the coordinates of the series
* `rwl`: the rwl object containing the single tree ring data

For plotting a shapefile of Europe is also included

```{r}
load("euro_shape.RData")
```

## Specific R functions

To better streamline the analysis workflow we coded a set of functions which are included in the `additional_functions.R` file. In particular:

* `calculate_matching`: This function efficiently calculates matching metrics between different series organized in a rwl object. 
* `averager`: uses the `chron` function of `dplR` to calculate an *average* rwl object from a sett of tree ring chronologies
* `log10_minor_break` : allows to add log spaced grid to a ggplot object


```{r}
source("additional_functions.R")
```


## Series pre-processing

The objective of this section is:

* detrend and combine the individual tree ring data in a coherent way
* assess the dataset distribution in terms of replication, length of the series, species distribution

Let's start with an initial assessment of the size of the dataset. We are dealing here with

```{r}
nrow(dat)
```

series with the following species distribution

```{r}
table(dat$specie)
```

In terms of spatial coverage at the continental scale 


```{r fig.height=3, fig.width=8}
geo_area %>% 
  ggplot() + 
  geom_sf() + 
  geom_sf(data = dat, aes(col = specie), size = 1, alpha = 0.5) + 
  scale_color_brewer(palette = "Set1", name = "Specie") + 
  facet_wrap(~specie) + 
  theme_bw() + 
  theme(legend.position = "none")

```

We also calculate the total number of tree ring series we are dealing with

```{r}
## add a column withe the samples per series
dat <- dat %>% 
  mutate(nsamp = map_int(rwl, ~ncol(.x))) 

## calculate the total number of tree ring series
sum(dat$nsamp)
```

```{r fig.height=3, fig.width=8}
dat %>% 
  ggplot() + 
  geom_histogram(aes(x = nsamp, fill = specie), col  ="white", alpha = 0.7, bins  = 10) + 
  scale_fill_brewer(palette = "Set1", name = "Specie") + 
  geom_vline(xintercept = 20, col = "red", lty = 2) +
  facet_wrap(~specie, scales = "free_y") + 
  theme_bw() + 
  scale_x_log10(minor_breaks=log10_minor_break()) + 
  xlab("Log (number of samples)") +
  ylab("No of site chronologies") +
  theme(aspect.ratio = 0.5)


```

So we see that on average we are dealing with around 25 samples per series. Here a table with a more complete summary of the series statistics 

```{r}
dat %>% 
  group_by(specie) %>% 
  st_drop_geometry() %>% 
  summarise(min_samp = min(nsamp),
            max_samp = max(nsamp),
            mean_nsamp = mean(nsamp)) %>% 
  kable("pipe")
  
```

We now detrend all the individual tree series by a Friedman supersmoother and then we calculate the average chronology of the detrended series.


```{r include=FALSE}
dat <- dat %>% 
  mutate(det_rwl = map(rwl, ~ as.rwl(detrend(.x, method = "Friedman")))) %>% 
  mutate(mean_chron_d = map2(study_code,det_rwl, ~ averager(.y,id = .x))) 
```

We have a limited numbers of warnings 22 out of 12000 not a lot I would say. The problem is in general due to the presence of bumpy trends that cannot be efficiently eliminated.

And we arrange everything into two big master objects

```{r}
master_d <- combine.rwl(dat$mean_chron_d)
```

Can we say something about the length of the master?

```{r fig.height=3, fig.width=8}
master_d %>% 
  rownames_to_column("year") %>% 
  mutate(year = as.numeric(year)) %>% 
  mutate(across(-year, ~ as.numeric(!is.na(.x)))) %>% 
  pivot_longer(-year, names_to = "study_code") %>% 
  filter(value > 0) %>% 
  group_by(study_code) %>% 
  summarise(serie_length = length(year)) %>% 
  left_join(dat %>% select(study_code, specie)) %>% 
  ggplot() + 
  geom_histogram(aes(x = serie_length, fill = specie), col  ="white", alpha = 0.7, bins  = 15) + 
  scale_fill_brewer(palette = "Set1", name = "Specie") + 
  facet_wrap(~specie, scales = "free_y") + 
  theme_bw() + 
  xlab("Length of the series (years)") + 
  ylab("No of site chronologies") +
  theme(aspect.ratio = 0.5)

```

As last step of exploratory analyssi we shoe the full extent of the overall dataset



```{r}
master_d %>% 
  rownames_to_column("year") %>% 
  mutate(year = as.numeric(year)) %>% 
  mutate(across(-year, ~ as.numeric(!is.na(.x)))) %>% 
  pivot_longer(-year, names_to = "study_code") %>% 
  filter(value > 0) %>% 
  group_by(study_code) %>% 
  summarise(serie_length = length(year)) %>% 
  left_join(dat %>% select(study_code, specie)) %>% 
  group_by(specie) %>% 
  summarise(min_length = min(serie_length),
            max_length = max(serie_length),
            mean_length = mean(serie_length)) %>% 
  kable("pipe")
```

We now need a plot to show the full set of series ...

```{r}
master_d %>% 
  rownames_to_column("year") %>% 
  mutate(year = as.numeric(year)) %>% 
  mutate(across(-year, ~ as.numeric(!is.na(.x)))) %>% 
  pivot_longer(-year, names_to = "study_code") %>% 
  filter(value > 0) %>% 
  group_by(study_code) %>% 
  summarise(y_s = min(year), y_m = max(year)) %>% 
  left_join(dat %>% select(study_code, specie)) %>%
  arrange(y_s) %>% 
  mutate(study_code = factor(study_code, levels = study_code)) %>% 
  ggplot() + 
  geom_segment(aes(x = y_s, xend = y_m, y = study_code, yend = study_code, col = specie), lwd = 0.5) + 
  coord_flip() + 
  scale_color_brewer(palette = "Set1", name = "Specie") + 
  theme_bw() + 
  xlab("Year") + 
  ylab("Series") +
   theme(axis.text.x=element_blank(),
         axis.ticks.y=element_blank(),
        aspect.ratio = 0.3
        )

  
  
```

## Correlation analysis of the series

Here we calculate the matching among all masters, to be conservative we consider only couples which are superimposing more than 50 years

```{r}
mtc_d <- calculate_matching(master_d, sup = 50)
```

```{r}
head(mtc_d) %>% 
  kable("pipe")
```

To investigate the association between similarity and distance we also calculate the physical distance between the set of points


```{r}
dist_mat <-  dat%>% 
  st_distance()/1000
rownames(dist_mat) <- dat$study_code
colnames(dist_mat) <- dat$study_code
```


We now combine the spatial and dendrochronological similarity, also keeping track of the specied of the couples which is taken from the `dat` tibble

```{r}
mtc_d <- mtc_d %>% 
  separate(series, into = c("s1","s2"), sep = "__") %>% 
  na.omit() %>% 
  mutate(d = map2_dbl(s1,s2, ~ dist_mat[.x,.y])) %>% 
  left_join(dat%>% st_drop_geometry() %>% 
              select(study_code,specie), by = c("s1" = "study_code")) %>% 
  left_join(dat %>% st_drop_geometry() %>% 
              select(study_code,specie), by = c("s2" = "study_code")) %>% 
  mutate(couple_id = case_when(specie.x == "LADE" & 
                                 specie.y == "LADE" ~ "L-L",
                               specie.x == "PCAB" & 
                                 specie.y == "PCAB" ~ "P-P",
                               specie.x == "ABAL" & 
                                 specie.y == "ABAL" ~ "A-A",
                               specie.x == "LADE" & 
                                 specie.y == "PCAB" ~ "L-P",
                               specie.y == "LADE" & 
                                 specie.x == "PCAB" ~ "L-P",
                               specie.x == "LADE" & 
                                 specie.y == "ABAL" ~ "L-A",
                               specie.y == "LADE" & 
                                 specie.x == "ABAL" ~ "L-A",
                               specie.x == "PCAB" & 
                                 specie.y == "ABAL" ~ "P-A",
                               specie.y == "PCAB" & 
                                 specie.x == "ABAL" ~ "P-A",
                              
                               )) 
  
```

And now we make a plot organized by specie

```{r fig.height=5, fig.width=7}
mtc_d %>% 
  filter(d > 0) %>% 
  mutate(couple_id = factor(couple_id, 
                            levels = c("A-A","L-A","L-L","L-P","P-P","P-A"))) %>% 
  #mutate(length_tile = factor(cut(lng,2))) %>% 
  ggplot() + 
  geom_hline(yintercept = 0, col  ="red", lty = 2) + 
  geom_point(aes(x = d, y = cor, col = lng), alpha = 0.9, pch = 1, size = 0.8) + 
  scale_color_binned_sequential(palette = "viridis", rev = FALSE, name = "Overlap (yr)") + 
  facet_wrap(~couple_id, ncol = 2) + 
  xlab("d (Km)") + 
  ylab("r (Pearson)") + 
  scale_x_log10(minor_breaks=log10_minor_break()) +
  theme_bw() + 
  theme(
        panel.grid.major=element_line (colour="gray80"), 
        panel.grid.minor=element_line(colour="gray80"),
        aspect.ratio = 0.3)

```


The plot shows the expected trend. The correlation increases at small distances and the extent of this association depends on the nature of the couple. In all cases, the vertical cloud @1e-01 shows the distribution of the correlation measured for masters assigned to the same spatial location.

Here we also propose a way to identify a reasonable distance starting from a measure of correlation. One idea is to fit a quantile regression line to the cloud of point which can be used to set an upper boundary to the correlation which can be measured for each distance.

We now focus on the homologous couples

```{r fig.height=3, fig.width=6}
mtc_d %>% 
  filter(d > 0) %>% 
  filter(couple_id %in% c("A-A","L-L","P-P")) %>% 
  ggplot() + 
  geom_hline(yintercept = 0, col  ="red", lty = 2) + 
  geom_point(aes(x = d, y = cor, col = couple_id), alpha = 0.8, pch = 1, cex = 0.8) + 
  #scale_shape_manual(values = c(1,3,5)) + 
  scale_color_discrete_qualitative(palette = "Dark2") + 
  xlab("d (Km)") + 
  ylab("r (Pearson)") + 
  scale_x_log10(minor_breaks=log10_minor_break()) +
  theme_bw() + 
  theme(
        panel.grid.major=element_line (colour="gray80"), 
        panel.grid.minor=element_line(colour="gray80"),
        aspect.ratio = 0.3)

```

## Quantile Regression

We apply a quantile regression approach to identify a 95% regression limit which can be used to estimate a distance from the correlation

```{r}
data_full_qr <- mtc_d %>% 
  filter(d > 0) %>% 
  filter(couple_id %in% c("A-A","L-L","P-P"))

full_quantreg <- rq(cor ~ log10(d), tau = 0.95, data = data_full_qr)

summ(full_quantreg, confint = TRUE)
```

The following plot shows the quantile regression line and the arrows illustrate how the line can be used to estimate the distance from a correlation value. As discussed in the paper, suppose that the correlation between an unknown series and a reference one turns out to be 0.75. The figure indicates that this level of similarity is unlikely to be found for distances larger than 7 km, so this distance can be used to perform an approximate dendroprovenance. 

```{r}
mtc_d %>% 
  filter(d > 0) %>% 
  filter(couple_id %in% c("A-A","L-L","P-P")) %>% 
  bind_cols(predict(full_quantreg, data_full_qr, interval = "confidence")) %>% 
  ggplot() + 
  geom_hline(yintercept = 0, col  ="red") + 
  geom_point(aes(x = d, y = cor, col = couple_id, pch = couple_id), alpha = 0.5, size = 0.6) + 
  geom_line(aes(x = d, y = fit), col = "darkred") + 
  geom_line(aes(x = d, y = lower), col = "red", alpha =  0.6, lwd = 0.5) + 
  geom_line(aes(x = d, y = higher), col = "red", alpha =  0.6, lwd = 0.5) + 
  annotate("segment", x = 0.5, xend = 7, y = 0.75, yend = 0.75, arrow = arrow(length = unit(0.2, "cm"))) + 
  annotate("segment", x = 7, xend = 7, y = 0.75, yend = 0, arrow = arrow(length = unit(0.2, "cm"))) + 
  scale_shape_manual(values = c(1,3,4), name = "Couple id") + 
  scale_color_brewer(palette = "Set1", name = "Couple id") + 
  xlab("d (Km)") + 
  ylab("r (Pearson)") + 
  scale_x_log10(minor_breaks=log10_minor_break()) +
  theme_bw() + 
  theme(
        panel.grid.major=element_line (colour="gray80"), 
        panel.grid.minor=element_line(colour="gray80"),
        aspect.ratio = 0.3)

```

Here we use the `broom` package to extract the coeeficients form the quantile regression object

```{r}
tidy(full_quantreg) %>% 
  kable("pipe")
```


We have shown how to construct the model for the three species together, we now run the analysis on the three separate groups of samples (L-L, P-P, A-A)
at two different quantiles:  0.95, 0.90

```{r}
quantregs <- mtc_d %>% 
  filter(d > 0) %>% 
  filter(couple_id %in% c("A-A","L-L","P-P")) %>% 
  nest(data = - couple_id) %>% 
  mutate(quantreg = map(data, ~ rq(cor ~ log10(d), tau = c(0.90,0.95), data = .x))) %>% 
  mutate(coefs = map(quantreg, ~ tidy(.x, conf.int = FALSE))) 
```


First we we prepare a plot with the coefficients

```{r}
quantregs %>% 
  select(couple_id,coefs) %>% 
  unnest(coefs) %>% 
  rename(c("quantile" = "tau")) %>% 
  mutate(quantile = factor(quantile)) %>% 
  mutate(term = ifelse(term == "(Intercept)","b","a")) %>%
  ggplot() + 
  geom_point(aes(x = couple_id, y = estimate, col = quantile, group = quantile), position = position_dodge(width = 0.5)) + 
  geom_linerange(aes(x = couple_id, ymin = conf.low, ymax = conf.high, col = quantile), position = position_dodge(width = 0.5)) + 
  facet_wrap(~term, scales = "free") +
  theme_bw() + 
  theme(
        panel.grid.major=element_line (colour="gray80"), 
        panel.grid.minor=element_line(colour="gray80"), 
        aspect.ratio = 1)
  

```

Which shows that the individual models yields comparable coefficients, the only exception being the intercept for P-P. As before, `broom` can be used to extract the coefficients 

```{r}
quantregs %>% 
  select(couple_id,coefs) %>% 
  unnest(coefs) %>% 
  rename(c("confidence" = "tau")) %>% 
  mutate(confidence = paste0(confidence*100," %")) %>% 
  mutate(term = ifelse(term == "(Intercept)","b","a")) %>% 
  kable("pipe")
```


## Role of elevation

Many of the series we have been collecting have also indications about the altitude of the samples and this parameter is known to affect the similarity. We arbitrarily set a threshold of 1500 meter to identify low/high altitude groups for the three species. The distribution of altitudes and the 1500m cutoff is shown in the following plot

```{r}

altitude_cutoff <- tibble(specie = c("PCAB","ABAL","LADE"), limit = c(1500,1500,1500))


dat %>% 
  na.omit() %>% 
  ggplot() + 
  geom_histogram(aes(x = maxelev, fill = specie), col = "white", alpha = 0.8,
                 position = position_identity()) + 
  geom_vline(data = altitude_cutoff, mapping = aes(xintercept = limit), col = "red", lty = 2) + 
  scale_fill_brewer(palette = "Set1", name = "Specie") + 
  facet_wrap(~specie, nrow = 1, scales = "free_y") + 
  xlab("Elevation") + 
  ylab("No of site chronologies") + 
  theme_bw() + 
  theme(aspect.ratio = 0.5)
  

```

We now split the series in in low/high altitude groups
```{r}
altitude_study <- dat %>% 
  st_drop_geometry() %>% 
  na.omit() %>% 
  select(specie,mean_chron_d,maxelev) %>% 
  nest(data = -specie) %>% 
  left_join(altitude_cutoff) %>% 
  mutate(high = map2(data, limit, function(x,y) {
    one <- x %>% filter(maxelev >= y)
    out <- combine.rwl(one$mean_chron_d)
    out
    })) %>% 
  mutate(low = map2(data, limit, function(x,y) {
    one <- x %>% filter(maxelev < y)
    out <- combine.rwl(one$mean_chron_d)
    out
    })) %>% 
  select(specie,high,low) %>% 
  pivot_longer(-specie, values_to = "rwl", names_to = "group")
  
  
  
```
And we calculate the similarity within each group

```{r}
altitude_study <- altitude_study %>% 
  mutate(mtc = map(rwl, ~ calculate_matching(.x, sup = 50)))
```

We also add the geographic distance

```{r}
altitude_study <- altitude_study %>% 
  mutate(mtc_full = map(mtc, ~ .x %>% 
  separate(series, into = c("s1","s2"), sep = "__") %>% 
  na.omit() %>% 
  mutate(d = map2_dbl(s1,s2, ~ dist_mat[.x,.y]))))
```

And now we finally plot the similarities

```{r fig.height=5, fig.width=7}
altitude_study %>% 
  select(specie, group, mtc_full) %>% 
  unnest(mtc_full) %>% 
  filter(d > 0) %>% 
  ggplot() + 
  geom_hline(yintercept = 0, col  ="red", lty = 2) + 
  geom_point(aes(x = d, y = cor, col = specie), size = 0.5, shape = 3) + 
  scale_color_brewer(palette = "Set1") + 
  facet_grid(specie~group) + 
  scale_x_log10(minor_breaks=log10_minor_break()) +
  scale_y_continuous(breaks=seq(-0.5,1,0.3)) + 
  xlab("d (Km)") + 
  ylab("r (Pearson)") + 
  theme_bw() + 
  theme(
        panel.grid.major=element_line (colour="gray80"), 
        panel.grid.minor=element_line(colour="gray80"),
        legend.position = "none",
        aspect.ratio = 0.3)

```



## Bibliography

-   Guiterman C. H., Gille E., Shepherd E., Mcneill S., Payne C. R., Morrill C., 2024. The International Tree-Ring Data Bank at Fifty: Status of Stewardship for Future Scientific Discovery. Tree-Ring Research, 80 (1): 13-18.
