---
title: "readme"
author: "Pietro Franceschi"
date: "`r Sys.Date()`"
output: github_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Introduction

This md file documents the data analysis workflow used to produce the results presented in two papers:

* <a href="https://www.esempio.com" target="_blank">*Correlation between tree-ring series as a dendroprovenancing evaluation tool*</a> Bernabei M., Franceschi P., 2024. Science of the Total Environment, 954: 176516

* M Bernabei and P. Franceschi *Reconsidering the use of t statistic in Dendroprovenancing*, in preparation


The first paper reports on an extensive study realized on a subset of the data available at the International Tree Ring Data Bank which investigate potential and the pitfalls of correlation analysis in dendroprovenancing. In the paper the authors propose the use of quantile regression to perform approximate dendroprovenancing starting from Pearson correlation values.

The second paper extends the proposed approach to the t-statistic, highlighting the effects of series overlap on this widely used statistic in dendrochronology for cross-dating and, at least in preliminary stages, for dendroprovenancing.



## Includes

This is the set of libraries required to perform the analyses

```{r message=FALSE, warning=FALSE}
library(dplR)                
library(tidyverse)
library(sf)
library(knitr)
library(magrittr)
library(quantreg)
library(colorspace)
library(jtools)
library(broom)
library(zoo)
```

## Data

11,972 individual, raw and georeferenced tree ring series, referring to three important Alpine species, Norway spruce (Picea abies Karst., PCAB), larch (Larix decidua Mill., LADE) and silver fir (Abies alba Mill., ABAL), were downloaded from the ITRDB site ([International Tree Ring Data Bank, United States’ National Oceanic and Atmospheric Administration Paleoclimatology Program and World Data Center for Paleoclimatology](https://www.ncei.noaa.gov/products/paleoclimatology/tree-ring) (Guiterman et al., 2024).

The data were organised into a complete tibble which is made available in the repository (`full_series.RData`).

```{r}
## load the dataset
load("full_series.RData")

## print the column names
colnames(dat)
```
* `meta`: the metainformation present in the NOAA db.
* `study_code`: the identifier of the study
* `specie`: the tree specie code
* `geotype`: the type of geographical object (here always a POINT)
* `minelev`,`maxelev`: the altitude boundaries of the series, if present
* `geometry`: the coordinates of the series
* `rwl`: the rwl object containing the single tree ring data

For plotting a shapefile of Europe is also included

```{r}
load("euro_shape.RData")
```

## Specific R functions

To better streamline the analysis workflow we coded a set of functions which are included in the `additional_functions.R` file. In particular:

* `calculate_matching`: This function efficiently calculates matching metrics between different series organized in a rwl object. 
* `averager`: uses the `chron` function of `dplR` to calculate an *average* rwl object from a sett of tree ring chronologies
* `log10_minor_break` : allows to add log spaced grid to a ggplot object
* `bp_transform`: apply the detrending included in the prcedure proposed by Baillie and Pilcher
* `holl_transform`: apply the detrending proposed by Hollstein


```{r}
source("additional_functions.R")
```


## Series pre-processing

The objective of this section is:

* detrend and combine the individual tree ring data in a coherent way exploring the effects of the different detrending
* assess the dataset distribution in terms of replication, length of the series, species distribution

Let's start with an initial assessment of the size of the dataset. We are dealing here with

```{r}
nrow(dat)
```

series with the following species distribution

```{r}
table(dat$specie)
```

In terms of spatial coverage at the continental scale:


```{r fig.height=3, fig.width=8}
geo_area %>% 
  ggplot() + 
  geom_sf() + 
  geom_sf(data = dat, aes(col = specie), size = 1, alpha = 0.5) + 
  scale_color_brewer(palette = "Set1", name = "Specie") + 
  facet_wrap(~specie) + 
  theme_bw() + 
  theme(legend.position = "none")

```

We also calculate the total number of tree ring series we are dealing with

```{r}
## add a column withe the samples per series
dat <- dat %>% 
  mutate(nsamp = map_int(rwl, ~ncol(.x))) 

## calculate the total number of tree ring series
sum(dat$nsamp)
```

```{r fig.height=3, fig.width=8}
dat %>% 
  ggplot() + 
  geom_histogram(aes(x = nsamp, fill = specie), col  ="white", alpha = 0.7, bins  = 10) + 
  scale_fill_brewer(palette = "Set1", name = "Specie") + 
  geom_vline(xintercept = 20, col = "red", lty = 2) +
  facet_wrap(~specie, scales = "free_y") + 
  theme_bw() + 
  scale_x_log10(minor_breaks=log10_minor_break()) + 
  xlab("Log (number of samples)") +
  ylab("No of site chronologies") +
  theme(aspect.ratio = 0.5)

```

So we see that on average we are dealing with around 25 samples per series. Here a table with a more complete summary of the series statistics 

```{r}
dat %>% 
  group_by(specie) %>% 
  st_drop_geometry() %>% 
  summarise(min_samp = min(nsamp),
            max_samp = max(nsamp),
            mean_nsamp = mean(nsamp)) %>% 
  kable("pipe")
  
```



On the individual series we now apply the following pipeline:

* we detrend the individual series with four different methods (Baille and Pilcher, Hollstein, spline based and Friedman)
* we calculate the mean chronology for all series and all detrendings




```{r include=FALSE}


detrend_bp <- function(rwl){
  out <- apply(rwl, 2, function(c) bp_transform(c))
  out
}


detrend_hol <- function(rwl){
  out <- apply(rwl,2,function(c) holl_transform(c))
  out
}

## Here we apply four different detrends:
# * Friedman supersmoother
# * standard spline
# * baillie pichler
# * hollstein


dat <- dat %>% 
  mutate(fried_det = map(rwl, ~ as.rwl(detrend(.x, method = "Friedman")), 
                         .progress = list(
                          type = "iterator", 
                          format = "Calculating {cli::pb_bar} {cli::pb_percent}",
                          clear = TRUE))) %>% 
  mutate(spline_det = map(rwl, ~ as.rwl(detrend(.x,method = "Spline")), 
                         .progress = list(
                          type = "iterator", 
                          format = "Calculating {cli::pb_bar} {cli::pb_percent}",
                          clear = TRUE))) %>% 
  mutate(bp_det = map(rwl, ~ detrend_bp(.x), 
                         .progress = list(
                          type = "iterator", 
                          format = "Calculating {cli::pb_bar} {cli::pb_percent}",
                          clear = TRUE))) %>% 
  mutate(hol_det = map(rwl, ~ detrend_hol(.x), 
                         .progress = list(
                          type = "iterator", 
                          format = "Calculating {cli::pb_bar} {cli::pb_percent}",
                          clear = TRUE)))
  

## here we transform each object into an rwl
pb <- txtProgressBar(min = 1, max = length(dat$hol_det), style = 3)

for(i in 1:length(dat$hol_det)){
  dat$hol_det[[i]] <- as.rwl(dat$hol_det[[i]])
  dat$bp_det[[i]] <- as.rwl(dat$bp_det[[i]])
  setTxtProgressBar(pb, i)
}
close(pb)


```


We have a limited numbers of warnings, not a lot I would say. The problem is in general due to the presence of bumpy trends that cannot be efficiently eliminated.


We calculate the mean chronology

```{r include=FALSE}
dat <- dat %>% 
  mutate(mean_chron_fried = map2(study_code,fried_det, ~ averager(.y,id = .x), .progress = list(
                          type = "iterator", 
                          format = "Calculating {cli::pb_bar} {cli::pb_percent}",
                          clear = TRUE))) %>% 
  mutate(mean_chron_spline = map2(study_code,spline_det, ~ averager(.y,id = .x),.progress = list(
                          type = "iterator", 
                          format = "Calculating {cli::pb_bar} {cli::pb_percent}",
                          clear = TRUE))) %>% 
  mutate(mean_chron_bp = map2(study_code,bp_det, ~ averager(.y,id = .x),.progress = list(
                          type = "iterator", 
                          format = "Calculating {cli::pb_bar} {cli::pb_percent}",
                          clear = TRUE))) %>% 
  mutate(mean_chron_hol = map2(study_code,hol_det, ~ averager(.y,id = .x),.progress = list(
                          type = "iterator", 
                          format = "Calculating {cli::pb_bar} {cli::pb_percent}",
                          clear = TRUE))) %>% 
  mutate(mean_chron = map2(study_code,rwl, ~ averager(.y,id = .x),.progress = list(
                          type = "iterator", 
                          format = "Calculating {cli::pb_bar} {cli::pb_percent}",
                          clear = TRUE)))
  
```



And we arrange everything into tibble of masters

```{r}

master <- tibble(detrend = c("raw","bp","hol","spline","friedman"),
                 master = list(
                   combine.rwl(dat$mean_chron),
                   combine.rwl(dat$mean_chron_bp),
                   combine.rwl(dat$mean_chron_hol),
                   combine.rwl(dat$mean_chron_spline),
                   combine.rwl(dat$mean_chron_fried)
                 ))
names(master$master) <- master$detrend

```


Can we say something about the length of the master?

```{r fig.height=3, fig.width=7}


master$master$bp %>% 
  rownames_to_column("year") %>% 
  mutate(year = as.numeric(year)) %>% 
  mutate(across(-year, ~ as.numeric(!is.na(.x)))) %>% 
  pivot_longer(-year, names_to = "study_code") %>% 
  filter(value > 0) %>% 
  group_by(study_code) %>% 
  summarise(serie_length = length(year)) %>% 
  left_join(dat %>% select(study_code, specie)) %>% 
  ggplot() + 
  geom_histogram(aes(x = serie_length, fill = specie), col  ="white", alpha = 0.7, bins  = 15) + 
  scale_fill_brewer(palette = "Set1", name = "Specie") + 
  facet_wrap(~specie, scales = "free_y") + 
  theme_bw() + 
  #scale_x_log10(minor_breaks=log10_minor_break()) + 
  xlab("Length of the series (years)") + 
  ylab("No of site chronologies") +
  theme(aspect.ratio = 0.5)


```
It is nice to extract a table with the characteristics of the dataset


```{r}
master$master$friedman %>% 
  rownames_to_column("year") %>% 
  mutate(year = as.numeric(year)) %>% 
  mutate(across(-year, ~ as.numeric(!is.na(.x)))) %>% 
  pivot_longer(-year, names_to = "study_code") %>% 
  filter(value > 0) %>% 
  group_by(study_code) %>% 
  summarise(serie_length = length(year)) %>% 
  left_join(dat %>% select(study_code, specie)) %>% 
  group_by(specie) %>% 
  summarise(min_length = min(serie_length),
            max_length = max(serie_length),
            mean_length = mean(serie_length)) %>% 
  kable("pipe")
```

We now need a plot to show the full set of series ...

```{r}
master$master$friedman %>% 
  rownames_to_column("year") %>% 
  mutate(year = as.numeric(year)) %>% 
  mutate(across(-year, ~ as.numeric(!is.na(.x)))) %>% 
  pivot_longer(-year, names_to = "study_code") %>% 
  filter(value > 0) %>% 
  group_by(study_code) %>% 
  summarise(y_s = min(year), y_m = max(year)) %>% 
  left_join(dat %>% select(study_code, specie)) %>%
  arrange(y_s) %>% 
  mutate(study_code = factor(study_code, levels = study_code)) %>% 
  ggplot() + 
  geom_segment(aes(x = y_s, xend = y_m, y = study_code, yend = study_code, col = specie), lwd = 0.5) + 
  coord_flip() + 
  scale_color_brewer(palette = "Set1", name = "Specie") + 
  theme_bw() + 
  xlab("Year") + 
  ylab("Series") +
   theme(axis.text.x=element_blank(),
         axis.ticks.y=element_blank(),
        aspect.ratio = 0.3
        )

  
  
```

## Correlation analysis of the series

Here we calculate the matching among all masters, to be conservative we consider only couples which are superimposing more than 50 years. Since we have five sets of masters associated to different detrending methods this is quite a slow task ... 

```{r}

master <- master %>% 
  mutate(matching = map(master, ~ calculate_matching(.x, sup = 50),
                        .progress = list(
                          type = "iterator", 
                          format = "Calculating {cli::pb_bar} {cli::pb_percent}",
                          clear = TRUE)))


```


A natural question that could be addressed using the full set of matching deals with the overall similarity of matches calculated with different detrending methods. This provides an indirect measure of the effects of detrending on correlation analysis. We can anticipate that these effects are expected to be significant.


```{r}
## first we find common couples
common_couples <- master %>% 
  select(detrend,matching) %>% 
  unnest(matching) %>% 
  na.omit() %>% 
  count(series) %>% 
  filter(n == 5) %>% 
  pull(series)
  
## then we transofr everything into a nice tibble
mtc_tibble <- master %>%
  select(detrend,matching) %>% 
  unnest(matching) %>% 
  filter(series %in% common_couples) %>% 
  pivot_longer(c(lng,cor)) %>% 
  mutate(name = paste0(detrend,"_",name)) %>% 
  select(-detrend) %>% 
  pivot_wider(names_from = name, values_from = value) 

head(mtc_tibble)
  
```

And now we are ready for a nice looking correlation plot 

```{r fig.height=6, fig.width=6}

mtc_tibble %>% 
  select(-ends_with("lng")) %>% 
  column_to_rownames("series") %>% 
  rename_with(.fn = function(nm) gsub("_cor","",nm)) %>% 
  as.matrix(.) %>% 
  pairs(., pch = ".")


```

This clearly indicate the effect of the choice of the detrending method on the series correlation matrix. The row signals are clearly uncorrelated with the detrended ones. Among the four types of detrending, Hollstein and Baille - Pilcher shows an high level of correlation. They also give results which are coherent with Friedman, while spline seems to yield a rather different pattern of similarity.

The previous plot highlights the effects of tetrending on the correlation across the full dataset, it would be interesting also to give a look to the effect of the detrending on the correlation one individual series. In other words: how similar is a serie with the detrended version of itself?


```{r}

## calculate the full set of intra series correlations
intra_serie_correlations <- master %>% 
  select(detrend,master) %>% 
  unnest_wider(master) %>% 
  pivot_longer(-detrend) %>% 
  nest(data = -name) %>% 
  mutate(dat_mat = map(data, function(d) {
    m <- do.call(cbind,d$value)
    m <- na.omit(m)
    colnames(m) <- d$detrend
    m
    
  })) %>% 
  select(-data)
  

```



```{r fig.height=6, fig.width=6}


reg <- function(x, y, ...) {
  points(x,y, ...)
  abline(lm(y~x)) 
  }# made to draw regression line instead of lowess line

panel.cor <- function(x, y, digits = 2, prefix = "", cex.cor, ...)
{
  usr_old <- par("usr")
  par(usr = c(0, 1, 0, 1))
  r <- abs(cor(x, y))
  txt <- format(c(r, 0.123456789), digits = digits)[1]
  txt <- paste0(prefix, txt)
  text(0.5, 0.5, txt, cex = 1, font = 1)
  par(usr = usr_old)
}
#EXAMPLE OF USE:
pairs(intra_serie_correlations$dat_mat[[5]], upper.panel = reg, # replace HERE for panel.smooth #
      cex = 1, pch = 19, col = adjustcolor(4, .4), cex.labels = 1, font.labels = 1, lower.panel = panel.cor)

```

The previous plot - obtained on a specific series - shows results which are slightly different with the overall pictures already discussed. Raw data are very much different from the detrended ones, but here Friedman and BP seems to give highly similar results while Hollstein and spline looks different.

Remember that in the previous plot each blue dot id the value of one year growth, while in the previous correlation plot each dot was the comparison between a couple of series.

In terms of "ecological" interpretation of these observations, it is not surprising that similarity depends on the detrending. Different choices will indeed remove time feature of different "frequency" highlighting the importance of short/medium/long scale ecological patterns on the similarity.

On this respect, however, it is important to highlight that for these reasons dendroprovenance could give more reliable results with a specific detrending method ... buth the best one cannot be decided "a priori".  

Just to visually demonstrate what we just discussed here we show aplot of the effects of setrending on a specific serie:

```{r}
intra_serie_correlations$dat_mat[[5]] %>% 
  as_tibble() %>% 
  rownames_to_column("index") %>% 
  mutate(index = as.numeric(index)) %>% 
  pivot_longer(-index, names_to = "Detrend") %>% 
  mutate(Detrend = factor(Detrend, levels = c("raw","bp","hol","friedman","spline"),
                          labels = c("Raw","Baillie-Pilcher","Hollstein","Friedman","Spline"))) %>% 
  ggplot() + 
  geom_line(aes(x = index, y = value, col = Detrend)) + 
  facet_wrap(~Detrend, ncol = 1, scales = "free") + 
  ylab("Amplitude (a.u.)") + 
  xlab("Years") + 
  theme_bw() + 
  theme(aspect.ratio = 0.1)
  
```

We clearly see:

* as expected raw data show large bumps
* spline detrending is not able to completely remove them
* the other three detrending strategies are able to remove multi year trends

## Association between series similarity and geographical distance

We now couple the physical distances to the correlation matrix

```{r}
dist_mat <-  dat%>% 
  st_distance() 
rownames(dist_mat) <- dat$study_code
colnames(dist_mat) <- dat$study_code


dist_tibble <- dist_mat %>% 
  as_tibble(rownames = "s1") %>% 
  pivot_longer(-s1, names_to = "s2", values_to = "d") %>% 
  mutate(series = paste0(s1,"__",s2)) %>% 
  mutate(d = as.vector(d)/1000) %>% 
  filter(d > 0.01) %>% left_join(dat%>% st_drop_geometry() %>% 
              select(study_code,specie), by = c("s1" = "study_code")) %>% 
  left_join(dat %>% st_drop_geometry() %>% 
              select(study_code,specie), by = c("s2" = "study_code")) %>% 
  mutate(couple_id = case_when(specie.x == "LADE" & 
                                 specie.y == "LADE" ~ "L-L",
                               specie.x == "PCAB" & 
                                 specie.y == "PCAB" ~ "P-P",
                               specie.x == "ABAL" & 
                                 specie.y == "ABAL" ~ "A-A",
                               specie.x == "LADE" & 
                                 specie.y == "PCAB" ~ "L-P",
                               specie.y == "LADE" & 
                                 specie.x == "PCAB" ~ "L-P",
                               specie.x == "LADE" & 
                                 specie.y == "ABAL" ~ "L-A",
                               specie.y == "LADE" & 
                                 specie.x == "ABAL" ~ "L-A",
                               specie.x == "PCAB" & 
                                 specie.y == "ABAL" ~ "P-A",
                               specie.y == "PCAB" & 
                                 specie.x == "ABAL" ~ "P-A",
                               )) 
  
  
## adding correlation similarity
dist_tibble <- dist_tibble %>% 
  right_join(mtc_tibble) %>% 
  mutate(min_overlap = pmin(raw_lng,bp_lng,hol_lng,spline_lng,friedman_lng)) %>% 
  select(-ends_with("lng"))
```

To show the complex interplay between the distance and the detrending we highlight in a plot the association between the correlation and the distance. To mak ethe plot clearer we bin the distance axis


```{r fig.height=6, fig.width=8}
dist_tibble %>% 
  pivot_longer(ends_with("cor")) %>% 
  filter(name != "raw_cor") %>% 
  na.omit() %>% 
  mutate(name = case_when(name == "bp_cor" ~ "Baillie-Pilcher",
                          name == "hol_cor" ~ "Hollstein",
                          name == "friedman_cor" ~ "Friedman",
                          name == "spline_cor" ~ "Spline")) %>% 
  mutate(couple_id = factor(couple_id, 
                            levels = c("A-A","L-A","L-L","L-P","P-P","P-A"))) %>% 
  mutate(dist_int = cut_interval(log10(d),6,dig.lab = 0)) %>% 
  group_by(dist_int,name, couple_id) %>% 
  summarise(median = median(value),
            q95 = quantile(value, 0.95),
            q5 = quantile(value,0.05)) %>% 
  #mutate(length_tile = factor(cut(lng,2))) %>% 
  ggplot() + 
  geom_hline(yintercept = 0, col  ="red", lty = 2) + 
  geom_point(aes(x = dist_int, y = median, col = name), alpha = 0.8, size = 2, position = position_dodge(width = 0.5)) + 
  geom_line(aes(x = dist_int, y = median, col = name, group = name), position = position_dodge(width = 0.5)) +
  geom_linerange(aes(x = dist_int, ymin = q5, ymax = q95, col = name),position = position_dodge(width = 0.5), alpha = 0.5) +
  scale_color_brewer(palette = "Set1", name = "Detrending") + 
  facet_wrap(~couple_id, ncol = 2) + 
  xlab("log10(d) (Km)") + 
  ylab("r (Pearson)") + 
  #scale_x_log10(minor_breaks=log10_minor_break()) +
  theme_bw() + 
  guides(color = guide_legend(override.aes = list(size = 3))) + 
  theme(
        panel.grid.major=element_line (colour="gray80"), 
        panel.grid.minor=element_line(colour="gray80"),
        axis.text.x = element_text(size = 8, angle = 45, hjust = 1))


```

Even if it is complex the previous plot is extremely informative. 

* In all cases correlation is negatively associated with geographic distance
* The strength of this association depends on the species we are comparing. e.g. on Larix - Larix the effect is stronger than on Larix - Abies
* Different detrending behave differently,a nd in general spline seems to be less effective in highlighting this trend.


The previous plot highlights general trends, but it cannot be directly used for dendroprovencance since it highlights "average" trends. To better pintpoint this aspect let's give a look to the full plot of the similarity between the series after Friedman detrending


```{r}
dist_tibble %>%  
  na.omit() %>% 
  mutate(couple_id = factor(couple_id, 
                            levels = c("A-A","L-A","L-L","L-P","P-P","P-A"))) %>% 
  #mutate(length_tile = factor(cut(lng,2))) %>% 
  ggplot() + 
  geom_hline(yintercept = 0, col  ="red", lty = 2) + 
  geom_point(aes(x = d, y = friedman_cor, col = min_overlap), alpha = 0.5, size = 0.8) + 
  scale_color_binned_sequential(palette = "viridis", rev = FALSE, name = "Overlap (yr)") + 
  facet_wrap(~couple_id, ncol = 2) + 
  xlab("d (Km)") + 
  ylab("r (Pearson)") + 
  scale_x_log10(minor_breaks=log10_minor_break()) +
  theme_bw() + 
  theme(
        panel.grid.major=element_line (colour="gray80"), 
        panel.grid.minor=element_line(colour="gray80"))

```


The plot shows the expected trend. The correlation increases at small distances and the extent of this association depends on the nature of the couple. 
What clearly stands out from the previous plot is that even if large correlations can be observed only at short distances, the reverse is not true: short distances can also be associated to "small" correlations. In other words, high correlations are always informative about distance, while low correlations are not. This observation makes perfectly sense and it is basically telling us: if you want to develop a quantitative model which uses correlation for dendroprovenanceing you should concentrate on modeling high correlations as a function of distance.

In order to do that our idea is to fit a quantile regression line to the cloud of points and this model will be useful to set an upper boundary to the correlation which can be measured for each distance.

Here we focus on the three homologous couples

```{r}
dist_tibble %>%  
  na.omit() %>% 
  filter(couple_id %in% c("A-A","L-L","P-P")) %>% 
  ggplot() + 
  geom_hline(yintercept = 0, col  ="red", lty = 2) + 
  geom_point(aes(x = d, y = friedman_cor, col = couple_id), alpha = 0.8, pch = 1, cex = 0.8) + 
  #scale_shape_manual(values = c(1,3,5)) + 
  scale_color_discrete_qualitative(palette = "Dark2") + 
  xlab("d (Km)") + 
  ylab("r (Pearson)") + 
  scale_x_log10(minor_breaks=log10_minor_break()) +
  theme_bw() + 
  theme(
        panel.grid.major=element_line (colour="gray80"), 
        panel.grid.minor=element_line(colour="gray80"), 
        aspect.ratio = 0.3)


```

We apply a quantile regression approach to identify a 95% regression limit which can be used to estimate a distance from the correlation. We do it putting the species together and as a function of the type of detrending

We want confidence intervals. So we use bootstrap on the data to do that we use the jtool package and we calculate bootstrap confidence intervals for the model coefficients

```{r}

full_qr <- dist_tibble %>%  
  na.omit() %>% 
  filter(couple_id %in% c("A-A","L-L","P-P")) %>% 
  select(d, ends_with("cor")) %>% 
  pivot_longer(ends_with("cor")) %>% 
  nest(data = -name) %>% 
  mutate(quantreg = map(data,~ rq(value ~ log10(d), tau = c(0.95,0.9), data = .x), 
                        .progress = list(
                          type = "iterator", 
                          format = "Calculating {cli::pb_bar} {cli::pb_percent}",
                          clear = TRUE))) %>% 
  mutate(coefs = map(quantreg, ~ tidy(.x, conf.int = TRUE, se.type = "boot"), .progress = list(
                          type = "iterator", 
                          format = "Calculating {cli::pb_bar} {cli::pb_percent}",
                          clear = TRUE)))


```


The following plot shows the quantile regression line and the arrows illustrate how the line can be used to estimate the distance from a correlation value. As discussed in the paper, suppose that the correlation between an unknown series and a reference one turns out to be 0.75. The figure indicates that this level of similarity is unlikely to be found for distances larger than 7 km, so this distance can be used to perform an approximate dendroprovenance. 

```{r}
dist_tibble %>%  
  na.omit() %>% 
  filter(d > 0) %>% 
  filter(couple_id %in% c("A-A","L-L","P-P")) %>% 
  bind_cols(predict(full_qr$quantreg[[5]], full_qr$data[[5]])) %>% 
  ggplot() + 
  geom_hline(yintercept = 0, col  ="red") + 
  geom_point(aes(x = d, y = friedman_cor, col = couple_id, pch = couple_id), alpha = 0.5, size = 0.6) + 
  geom_line(aes(x = d, y = `tau= 0.95`), col = "darkred") + 
  #geom_line(aes(x = d, y = lower), col = "red", alpha =  0.6, lwd = 0.5) + 
  #geom_line(aes(x = d, y = higher), col = "red", alpha =  0.6, lwd = 0.5) + 
  geom_segment(aes(x = 0.5, xend = 7, y = 0.75, yend = 0.75), arrow = arrow(length = unit(0.2, "cm"))) + 
  geom_segment(aes(x = 7, xend = 7, y = 0.75, yend = 0), arrow = arrow(length = unit(0.2, "cm"))) + 
  scale_shape_manual(values = c(1,3,4), name = "Couple id") + 
  scale_color_brewer(palette = "Set1", name = "Couple id") + 
  xlab("d (Km)") + 
  ylab("r (Pearson)") + 
  scale_x_log10(minor_breaks=log10_minor_break()) +
  theme_bw() + 
  theme(
        panel.grid.major=element_line (colour="gray80"), 
        panel.grid.minor=element_line(colour="gray80"), 
        aspect.ratio = 0.3)

```

We now want to have nice outputs for each one of the couples, and create a nice looking table and plot for the supplementary




```{r}
table_full_qr <- full_qr %>%
  select(name, coefs) %>%
  unnest(coefs) %>%
  rename("Detrending" = "name") %>%
  mutate(
    Detrending = case_when(
      Detrending == "bp_cor" ~ "Baillie-Pilcher",
      Detrending == "hol_cor" ~ "Hollstein",
      Detrending == "friedman_cor" ~ "Friedman",
      Detrending == "spline_cor" ~ "Spline",
      Detrending == "raw_cor" ~ "no detrend"
    )
  ) %>%
  select(-c("statistic", "p.value", "std.error")) %>%
  mutate(term = case_when(term == "(Intercept)" ~ "b", term == "log10(d)" ~ "a")) %>%
  rename(
    "Term" = "term",
    "Conf.low" = "conf.low",
    "Conf.high" = "conf.high",
    "Values" = "estimate",
    "Quantile" = "tau"
  ) %>%
  mutate(across(-c("Term","Detrending"),  ~ round(.x, 2))) 

table_full_qr %>% 
  kable("pipe")
  
```

To facilitate the comparison the previous data can be also visualized in graphical form:


```{r fig.height=5, fig.width=8}
full_qr %>%
  select(name, coefs) %>%
  unnest(coefs) %>%
  rename("Detrending" = "name") %>%
  mutate(
    Detrending = case_when(
      Detrending == "pb_cor" ~ "BP",
      Detrending == "hol_cor" ~ "Hollstein",
      Detrending == "friedman_cor" ~ "Friedman",
      Detrending == "spline_cor" ~ "Spline",
      Detrending == "raw_cor" ~ "no detrend"
    )
  ) %>%
  select(-c("statistic", "p.value", "std.error")) %>%
  mutate(term = case_when(term == "(Intercept)" ~ "b", term == "log10(d)" ~ "a")) %>% 
  rename("Quantile" = "tau") %>% 
  mutate(Quantile = factor(Quantile)) %>% 
  ggplot() + 
  geom_point(aes(x = Detrending, y = estimate, col = Quantile), position = position_dodge(width = 0.5)) + 
  geom_linerange(aes(x = Detrending, ymin = `conf.low`, ymax = `conf.high`, col = Quantile), position = position_dodge(width = 0.5)) + 
  facet_wrap(~term, scales = "free") + 
  xlab("") + 
  theme_bw() + 
  theme(aspect.ratio = 0.7,
        panel.grid.major.x = element_line(colour = "gray50", linetype = 2))
```


For practical applications we now calculate the quantile regression for the individual species by using a Friedman detrend as discussed in the STOTEN paper


```{r}

fried_qr <- dist_tibble %>%  
  na.omit() %>% 
  filter(couple_id %in% c("A-A","L-L","P-P")) %>% 
  select(d, couple_id,friedman_cor) %>% 
  nest(data = -couple_id) %>% 
  mutate(quantreg = map(data,~ rq(friedman_cor ~ log10(d), tau = c(0.95,0.9), data = .x), 
                        .progress = list(
                          type = "iterator", 
                          format = "Calculating {cli::pb_bar} {cli::pb_percent}",
                          clear = TRUE))) %>% 
  mutate(coefs = map(quantreg, ~ tidy(.x, conf.int = TRUE, se.type = "boot"), .progress = list(
                          type = "iterator", 
                          format = "Calculating {cli::pb_bar} {cli::pb_percent}",
                          clear = TRUE)))


```


And we supparise the outputs in a table


```{r}
fried_qr %>%
  select(couple_id, coefs) %>%
  unnest(coefs) %>%
  select(-c("statistic", "p.value", "std.error")) %>%
  mutate(term = case_when(term == "(Intercept)" ~ "b", term == "log10(d)" ~ "a")) %>%
  rename(
    "Term" = "term",
    "Conf.low" = "conf.low",
    "Conf.high" = "conf.high",
    "Values" = "estimate",
    "Quantile" = "tau",
    "Couple_id" = "couple_id"
  ) %>%
  mutate(across(-c("Term", "Couple_id"),  ~ round(.x, 2))) %>%
  kable("pipe")
  
 
```

## Role of elevation

Altitude difference is known to affect the correlation between series. In this section we present some considerations on this aspect which can be drawn because many of the series we have been collecting have also indications about the altitude of the samples.


```{r fig.height=3, fig.width=6}

## manually set limits to divade series into high and low altitude
altitude_cutoff <- tibble(specie = c("PCAB","ABAL","LADE"), limit = c(1500,1500,1500))


dat %>% 
  na.omit() %>% 
  ggplot() + 
  geom_histogram(aes(x = maxelev, fill = specie), col = "white", alpha = 0.8,
                 position = position_identity()) + 
  geom_vline(data = altitude_cutoff, mapping = aes(xintercept = limit), col = "red", lty = 2) + 
  scale_fill_brewer(palette = "Set1", name = "Specie") + 
  facet_wrap(~specie, nrow = 1, scales = "free_y") + 
  xlab("Elevation") + 
  ylab("No of site chronologies") + 
  theme_bw() +
  theme(aspect.ratio = 0.5)



```

As expected the three species are showing different altitude distributions. In the case of Larix the majority of the series was collected at relatively high altitudes, while the situation is reversed as far as Abies is concerned. The situation for Picea is somehow mixed: also in this case the majority of the samples were collected at low altitude even if the number of series measured above 1500 m is not negligible

Let's now look to the distance-correlation splitting the series in two altitude groups focusing on the homologous series:

```{r}
altitude_study <- dat %>% 
  st_drop_geometry() %>% 
  na.omit() %>% 
  select(specie,mean_chron_fried,maxelev) %>% 
  nest(data = -specie) %>% 
  left_join(altitude_cutoff) %>% 
  mutate(high = map2(data, limit, function(x,y) {
    one <- x %>% filter(maxelev >= y)
    out <- combine.rwl(one$mean_chron_fried)
    out
    })) %>% 
  mutate(low = map2(data, limit, function(x,y) {
    one <- x %>% filter(maxelev < y)
    out <- combine.rwl(one$mean_chron_fried)
    out
    })) %>% 
  select(specie,high,low) %>% 
  pivot_longer(-specie, values_to = "rwl", names_to = "group")
  
  
## calculate the similarity  
altitude_study <- altitude_study %>% 
  mutate(mtc = map(rwl, ~ calculate_matching(.x, sup = 50)))

altitude_study <- altitude_study %>% 
  mutate(mtc_full = map(mtc, ~ .x %>% 
  separate(series, into = c("s1","s2"), sep = "__") %>% 
  na.omit() %>% 
  mutate(d = map2_dbl(s1,s2, ~ dist_mat[.x,.y]))))

altitude_study %>% 
  select(specie, group, mtc_full) %>% 
  unnest(mtc_full) %>% 
  filter(d > 0) %>% 
  #mutate(couple_id = factor(couple_id, 
  #                          levels = c("P-P","L-L","P-L"))) %>% 
  #mutate(length_tile = factor(cut(lng,2))) %>% 
  ggplot() + 
  geom_hline(yintercept = 0, col  ="red", lty = 2) + 
  geom_point(aes(x = d, y = cor, col = specie), size = 0.5, shape = 3) + 
  scale_color_brewer(palette = "Set1") + 
  facet_grid(specie~group) + 
  scale_x_log10(minor_breaks=log10_minor_break()) +
  scale_y_continuous(breaks=seq(-0.5,1,0.3)) + 
  xlab("d (Km)") + 
  ylab("r (Pearson)") + 
  theme_bw() + 
  theme(
        panel.grid.major=element_line (colour="gray80"), 
        panel.grid.minor=element_line(colour="gray80"),
        legend.position = "none")


```
The previous plot shows the expected splitting of the species which makes difficult to draw conclusions for Larix and Abies. In the case of Picea the plot suggests the presence of a larger fraction of low correlations in the low altitude group, while the high correlation trends seems to be comparable even if the number of samples at low distances is unfortunately too small to draw robust conclusions. 



## From r to t

T statistic is often used for dendroproveneance, the fundamental difference between *t* and *r* arises from the fact that the t statistic also take into account the overlap between the series. This fact makes t more "reliable" as measure of similarity because its value tends to increase as the overall overlap increases. This fact somehow formalizes the reasonable observation that high similarities between two series are more trustworthy if the similarity extend over many years.

The mathematical between *t* and *s* is the following:

$$
t = \frac{r \sqrt{n - 2}}{\sqrt{1 - r^2}}
$$


In the previous sections we proposed to use quantile regression to predict r from d, so our results could be extended to link d and t considering different scenarios for n. What we would highlight about the formula is its non linearity in *r* and *n*. This implies that the quantile regression line will be transformed into a non linear two dimensional surface. Which in the following will be represented as a set of iso-n curves.

Let's start from the visualization of the trends for the quantile regression line constructed on the data of the three conifers taken together:


```{r}

## dataframe fro the prediction
newdata_full <- tibble(d = c(seq(1,9,1),seq(10,90,10),seq(100,1000,100)))
bp_qr <- full_qr$quantreg[[2]]

## prediction or the t-d relation for overlaps of 50, 100 and 200 years
newdata_full <- newdata_full %>% 
  add_column(r = predict(bp_qr, newdata_full)[,1]) %>% 
  mutate(t_50 = (r*sqrt(50 - 2))/sqrt(1-r^2)) %>% 
  mutate(t_100 = (r*sqrt(100 - 2))/sqrt(1-r^2)) %>% 
  mutate(t_200 = (r*sqrt(200 - 2))/sqrt(1-r^2))


newdata_full %>% 
  select(-r) %>% 
  pivot_longer(starts_with("t")) %>% 
  mutate(name = factor(name, levels = c("t_50","t_100","t_200"))) %>% 
  ggplot() + 
   geom_hline(aes(yintercept = 7), lty = 1) + 
  geom_point(aes(x = d, y = value, col = name)) +
  geom_line(aes(x = d, y = value, col = name)) + 
  scale_color_brewer(palette = "Set1", name = "Overlap (yr)") + 
  scale_x_log10() +
  xlab("d (Km)") + 
  ylab("t-statistics") + 
  scale_x_log10(minor_breaks=log10_minor_break()) +
  theme_bw() + 
  theme(
        panel.grid.major=element_line (colour="gray80"), 
        panel.grid.minor=element_line(colour="gray80"), 
        aspect.ratio = 0.5)
```

The non linear trend is clear. The same plot also shows the overall increase of the d-t curve as the number of samples increases. This is in keeping with the added confidence connected to a longer period in overlap.

As in the case of the linear relationship discussed above, the intersection between the horizontal line (corresponding to a t of 7) and the d-t curves shows the estimated maximal separation of two series showing a t of 7.

What should be observed here is that the same level of "similarity" (here 7) returns a different estimate of the maximal distance which grows as the overlap grows (10km for 50 years, 50 km fro 100 years, 150 km for 200 years). At first sight this result would seem strange, but in reality it fits with what we expect. In presence of large overlaps a t of 7 is not informative because we expect more than that if the two series are really similar.

For practical applications this means that the choice of an "informative" t for dendroprovenencing is not absolute but should take into consideration also the series overlap.

At a more subtle inspection, the previous plot also indicate that longer series should be always preferred. The difference on the ts obtained at 1km and 1000 km is indeed growing as the overlap grows. This implies that the green curve is more steep and contrasted and this will result in a more accurate devinition of the spatial separation between two series.

We conclude this notes by developing the previous analysis for the three separate species. Since BP detrending coupled with t statistics is often used in dendrochronological application we perform this analysis on the BP detrended curves. 

```{r}

bp_qr_t <- dist_tibble %>%  
  na.omit() %>% 
  filter(couple_id %in% c("A-A","L-L","P-P")) %>% 
  select(d, couple_id,bp_cor) %>% 
  nest(data = -couple_id) %>% 
  mutate(quantreg = map(data,~ rq(bp_cor ~ log10(d), tau = c(0.95,0.9), data = .x), 
                        .progress = list(
                          type = "iterator", 
                          format = "Calculating {cli::pb_bar} {cli::pb_percent}",
                          clear = TRUE))) %>% 
  mutate(coefs = map(quantreg, ~ tidy(.x, conf.int = TRUE, se.type = "boot"), .progress = list(
                          type = "iterator", 
                          format = "Calculating {cli::pb_bar} {cli::pb_percent}",
                          clear = TRUE)))


```

We now add the predictions for the t to each model

```{r}
bp_qr_t <- bp_qr_t %>% 
  mutate(t_predictions = map(quantreg, function(x){
    
    newdata <- tibble(d = c(seq(1,9,1),seq(10,90,10),seq(100,1000,100)))
    newdata <- newdata %>% 
      add_column(r = predict(x, newdata)[,1]) %>% 
      mutate(t_50 = (r*sqrt(50 - 2))/sqrt(1-r^2)) %>% 
      mutate(t_100 = (r*sqrt(100 - 2))/sqrt(1-r^2)) %>% 
      mutate(t_200 = (r*sqrt(200 - 2))/sqrt(1-r^2))
    
    return(newdata)
    
  }))
```

Ok now we prepare a faceted plot with the three species

```{r fig.height=4, fig.width=10}
bp_qr_t %>% 
  mutate(couple_id = recode(couple_id, "A-A" = "ABAL", "L-L" = "LADE", "P-P" = "PCAB")) %>% 
  select(couple_id,t_predictions) %>% 
  unnest(t_predictions) %>% 
  select(-r) %>% 
  pivot_longer(starts_with("t")) %>% 
  mutate(name = factor(name, levels = c("t_50","t_100","t_200"))) %>% 
  filter(!is.nan(value)) %>% 
  ggplot() + 
  geom_hline(aes(yintercept = 7), lty = 1) + 
  #geom_point(aes(x = d, y = value, col = name)) +
  geom_line(aes(x = d, y = value, col = name)) + 
  facet_wrap(~couple_id, scales = "free") +
  scale_color_brewer(palette = "Set1", name = "Overlap (yr)") + 
  scale_x_log10() +
  xlab("d (Km)") + 
  ylab("t-statistics") + 
  scale_x_log10(minor_breaks=log10_minor_break()) +
  theme_bw() + 
  theme(
        panel.grid.major=element_line (colour="gray80"), 
        panel.grid.minor=element_line(colour="gray80"), 
        aspect.ratio = 0.5)
  
 
```


## Bibliography

-   Guiterman C. H., Gille E., Shepherd E., Mcneill S., Payne C. R., Morrill C., 2024. The International Tree-Ring Data Bank at Fifty: Status of Stewardship for Future Scientific Discovery. Tree-Ring Research, 80 (1): 13-18.
